---
title: "Analyzing health and economic effects of storms on US"
author: "Rahul Jain"
date: "24/01/2022"
output: html_document
---

# 1 Synopsis : 

The goal of the project is to explore the NOAA Storm Dataset and analyze the impact of storms on US economy and health.
The dataset covers the time prios between 1950 and November 2011.

Health Impact from Storms - Tornado is the biggest contributor to fatalities contributing nearly 100k injuries or deaths. Other events contribution small in comparison to tornadoes

Economic Impact from Storms - Floods and hurricanes are the biggest contributors to impact on Economy

# 2 Dataset Source and other details : 

Dataset Source : https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

Other useful information about dataset:

* National Weather Service Storm Data Documentation : (https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

* National Climatic Data Center Storm Events FAQ (https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

# 3 Importing the libraries

```{r echo=TRUE}
library(ggplot2)
library(reshape2)
```

# 4 Data Processing

## 4.1 Downloading the data and loading into R

```{r echo=TRUE,cache=TRUE}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
download.file(fileUrl, destfile = paste0("/Users/jnrah/Downloads", '/repdata%2Fdata%2FStormData.csv.bz2'))
stormDF <- read.csv("/Users/jnrah/Downloads/repdata%2Fdata%2FStormData.csv.bz2",na.strings = c(""))
```

## 4.2 Data Exploration

### 4.2.1 Looking at the structure and summary of the dataset

```{r echo=TRUE}
head(stormDF)
```

```{r echo=TRUE}
str(stormDF)
```

```{r echo=TRUE}
summary(stormDF)
```

### 4.2.2 Looking at the amount of missing values in the data

```{r echo=TRUE}
sapply(stormDF,function(y) sum(length(which(is.na(y)))))
```

There are 37 variables in the data. Since we only need to examine health and economic impacts of storms on US, we
can remove the variables which are not required for the analysis.

## 4.3 Data Subsetting

### 4.3.1 Only keeping variables useful for analysis

```{r echo=TRUE}
col2keep <- c('BGN_DATE','EVTYPE','FATALITIES','INJURIES','PROPDMG','PROPDMGEXP','CROPDMG','CROPDMGEXP')
stormDF_sub <- stormDF[,col2keep]
stormDF_sub <- subset(stormDF_sub,EVTYPE != "?")
stormDF_sub$BGN_DATE <- as.Date(stormDF_sub$BGN_DATE,"%m/%d/%Y")
```

### 4.3.2 Subsetting data to create 2 different dataframes, one for analyzing health impact and other for economic impact

```{r echo=TRUE}
stormDF_health <- stormDF[,c('BGN_DATE','EVTYPE','FATALITIES','INJURIES')]
stormDF_health <- subset(stormDF_health,FATALITIES > 0 | INJURIES > 0)
```

```{r echo=TRUE}
stormDF_eco <- stormDF[,c('BGN_DATE','EVTYPE','PROPDMG','PROPDMGEXP','CROPDMG','CROPDMGEXP')]
stormDF_eco <- subset(stormDF_eco,PROPDMG > 0 | CROPDMG > 0)
```

### 4.3.3 Converting columns in stormDF_eco from exponent to numeric values

```{r echo=TRUE}
cropDmgKey <- c("?" = 10^0,"0" = 10^0,"K"=10^3,"M"=10^6,"B"=10^9)
stormDF_eco$CROPDMGEXP <- as.character(stormDF_eco$CROPDMGEXP)
stormDF_eco$CROPDMGEXP <- toupper(stormDF_eco$CROPDMGEXP)
stormDF_eco$CROPDMGEXP <- cropDmgKey[stormDF_eco$CROPDMGEXP]
stormDF_eco$CROPDMGEXP[is.na(stormDF_eco$CROPDMGEXP)] <- 1
```

```{r echo=TRUE}
propDmgKey <- c("?" = 1,"-"=1,"+"=1,"0" = 1,"2"=10^2,"3"=10^3,"4"=10^4,"5"=10^5,
                "6"=10^6,"7"=10^7,"H"=10^6,
                "K"=10^3,"M"=10^6,"B"=10^9)
stormDF_eco$PROPDMGEXP <- as.character(stormDF_eco$PROPDMGEXP)
stormDF_eco$PROPDMGEXP <- toupper(stormDF_eco$PROPDMGEXP)
stormDF_eco$PROPDMGEXP <- propDmgKey[stormDF_eco$PROPDMGEXP]
stormDF_eco$PROPDMGEXP[is.na(stormDF_eco$PROPDMGEXP)] <- 1
```


### 4.3.4 Making Economic Cost Columns

```{r echo=TRUE}
stormDF_eco$prop.cost <- stormDF_eco$PROPDMGEXP * stormDF_eco$PROPDMG
stormDF_eco$crop.cost <- stormDF_eco$CROPDMGEXP * stormDF_eco$CROPDMG
stormDF_eco$total.cost <- stormDF_eco$prop.cost + stormDF_eco$crop.cost
```

### 4.3.5 Making Health Cost Columns

```{r echo=TRUE}
stormDF_health$total_fatal <- stormDF_health$FATALITIES + stormDF_health$INJURIES
```

## 4.4 Creating summary dataframes for visualization

### 4.4.1 Economic Cost Dataframe

```{r echo=TRUE}
totalCost <- aggregate(stormDF_eco$total.cost~stormDF_eco$EVTYPE,FUN = sum)
names(totalCost) <- c("Event.Type","Total.Economic.Cost")
```

### 4.4.2 Health Cost Dataframe

```{r echo=TRUE}
healthCost <- aggregate(stormDF_health$total_fatal~stormDF_health$EVTYPE,FUN = sum)
names(healthCost) <- c("Event.Type","Total.Health.Cost")
```

#5 Results

## 5.1 Analyzing Health Cost

```{r echo=TRUE}
healthCost <- healthCost[order(-healthCost$Total.Health.Cost),]
top10_healthCost <- head(healthCost,10)
g1 <- ggplot(top10_healthCost,aes(reorder(Event.Type,-Total.Health.Cost),Total.Health.Cost))
g1 + geom_bar(stat = "identity",aes(fill=Event.Type),position = "dodge") + ylab("Frequency Count")  +
    xlab("Event Type") + theme(axis.text.x = element_text(angle=45, hjust=1)) +
    ggtitle("Top 10 US Killers") + theme(plot.title = element_text(hjust = 0.5))
```

## 5.2 Analyzing Economic Cost

```{r echo=TRUE}
ecoCost <- totalCost[order(-totalCost$Total.Economic.Cost),]
top10_ecoCost <- head(ecoCost,10)
g2 <- ggplot(top10_ecoCost,aes(reorder(Event.Type,-Total.Economic.Cost),Total.Economic.Cost))
g2 + geom_bar(stat = "identity",aes(fill=Event.Type),position = "dodge") + ylab("Frequency Count")  +
    xlab("Event Type") + theme(axis.text.x = element_text(angle=45, hjust=1)) +
    ggtitle("Top 10 Events Causing harm to Economy") + theme(plot.title = element_text(hjust = 0.5))
```